---
title: "Functional Incident HTN"
author: "LF Yeo"
date: "2024-11-11"
output: html_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message=FALSE) 
```
```{r, echo = FALSE}
library(magrittr)
library(RNOmni)
library(dplyr)
library(readr)
library(mia)
devtools::load_all()
```
```{r}
#Defining covariates for the model
vars <- list(BL_AGE = "Age",
             MEN = "Men",
             BMI = "BMI",
             PREVAL_DIAB = "Diabetes",
             SYSTM = "Systolic blood pressure",
             DIASM = "Diastolic blood pressure",
             HFC = "Healthy Food Choices",
             BP_TREAT = "Antihypertensive medication",
             PREVAL_HFAIL_STRICT = "Heart failure",
             PREVAL_HIBP = "Prevalent hypertension code",
             INCIDENT_HIBP = "Incident hypertension code",
             HIBP_AGEDIFF = "Hypertension code time",
             PREVAL_RX_RR_SPECIFIC = "Prevalent antihypertensive",
             INCIDENT_RX_RR_SPECIFIC = "Incident antihypertensive",
             RX_RR_SPECIFIC_AGEDIFF = "Antihypertensive agediff",
             PREVAL_HTN = "Prevalent hypertension",
             INCIDENT_HTN = "Incident hypertension",
             HTN_AGEDIFF = "Hypertension time (yrs)",
             CURR_SMOKE = "Smoking",
             EAST = "Eastern Finland",
             shannon = "Shannon diversity",
             total_reads = "Total reads") 

#Importing metadata and cleaning up
#needed to match SampleID and covariates
tse <- readRDS("../data/tse_mgs-20241118_104759.rds") %>%
  mia::transformAssay(assay.type = "counts", method = "relabundance") %>% 
  mia::estimateDiversity(assay.type = "counts", index = "shannon", name = "shannon") %>%
  tse_add_food_score(HFC) %>%
  tse_mutate(PREVAL_HTN = ifelse(PREVAL_HIBP + PREVAL_RX_RR_SPECIFIC > 0, 1, 0)) %>%
  tse_mutate(INCIDENT_HTN = ifelse(INCIDENT_HIBP + INCIDENT_RX_RR_SPECIFIC > 0, 1, 0)) %>%
  tse_mutate(HTN_AGEDIFF = pmin(HIBP_AGEDIFF, RX_RR_SPECIFIC_AGEDIFF)) %>% 
  tse_mutate(dplyr::across(c(MEN,
                             EAST,
                             BP_TREAT,
                             CURR_SMOKE,
                             dplyr::contains("INCIDENT"),
                             dplyr::contains("PREVAL")), as.factor)) %>%
  tse_filter(GRAVID %in% c(1, NA), BL_USE_RX_J01_1mo %in% c(0, NA)) %>% 
  tse_filter(SYSTM < 140, DIASM < 90, PREVAL_HTN == 0, HYPERT_AHA == 0) %>%
  { if (args$east) tse_filter(., EAST == 0) else . } %>% 
  { if (args$west) tse_filter(., EAST == 1) else . } %>% 
  tse_filter(dplyr::if_all(dplyr::one_of(names(vars)), not_na)) %>%
  tse_filter(total_reads > 35000) %>%
  tse_select(PREVAL_HTN, INCIDENT_HTN, HTN_AGEDIFF, names(vars))
```
```{r}
#Functional pathways from metaCyc
joined_pathabundance <- read_delim("../data/joined_pathabundance.tsv", 
                                   delim = "\t", escape_double = FALSE, trim_ws = TRUE)
```
```{r}
#Matching sampleID from functional data to metadata
#rename 1st column name
a <- as.data.frame(colData(tse))%>%
  tibble::rownames_to_column("Barcode")

# Get the column names (excluding the first one, which is probably the path/feature column)
sample_names <- colnames(joined_pathabundance)[-1]

# Clean sample names by removing after .R1
cleaned_sample_names <- gsub("\\.R1.*","", sample_names, perl = TRUE)

# Update the column names with the cleaned sample names
colnames(joined_pathabundance)[-1] <- cleaned_sample_names

# Find the intersection of sample names between the cleaned joined_path_abundance sample names and FINRISK sample data
common_samples <- intersect(a$Barcode, cleaned_sample_names)

# Subset the joined_pathabundance file to include only the common samples
joined_pathabundance<- joined_pathabundance %>%
  select(all_of(c("Pathway", common_samples)))

```
Filtering any pathways associated w/ species
```{r}
#Analysis focused on finding species pathways
shit<- joined_pathabundance %>% 
  filter(rowSums(across(where(is.numeric)))!=0) %>%
  #grep all Pathways that end with species
  filter(grepl('.s__', Pathway))
```
Calculating coefficient of variance
```{r}
#calculate standard coefficient variation by row-wise feature
#shit %>%
 # rowwise() %>%
#  mutate(
 #   mean_val = mean(c_across(2:6567)),
  #  sd_val = sd(c_across(2:6567)),
   # cv = (sd_val / mean_val) * 100) %>%
#  ungroup() %>%
 # hist(cv)
```
Transpose dataframe
```{r}
shit.t <- as.data.frame(t(shit))
#take first row as colname
colnames(shit.t) <- shit.t[1,]
#remove extra row
shit.t <- shit.t[-1, ] 
#change from character to numeric
shit.t <- shit.t %>% mutate_if(is.character, as.numeric) %>%
  tibble::rownames_to_column("Barcode") 
```
```{r}
write_rds(shit.t, "path.abund.t.rds")
write_rds(shit, "path.abund.rds")
```
Prevalence filtering
```{r}
#You will get a dataframe of prevalent pathways
#poo needs to have samples as row, pathways as column
shit.10 <- shit.t %>% 
  select(-1) %>%
  #mutate to 1 for everything > 0.0000000001
  mutate(across(everything(), ~ . > 10^-10)) %>%
  #get proportion
  summarize(across(everything(), mean)) %>%
  tidyr::gather(Pathway,prevalence) %>%
  #filter for pathways that are prevalent in 10%
  dplyr::filter(prevalence >0.05) 

# filtered to only prevalent pathways
hi <- shit.10 %>% 
  select(-2) %>%
  dplyr::inner_join(shit, by="Pathway" ) %>%
  t() %>%
  as.data.frame()

#take first row as colname
colnames(hi) <- hi[1,]
#remove extra row
hi <- hi[-1, ] 
#change from character to numeric
hi <- hi %>% mutate_if(is.character, as.numeric) %>%
  tibble::rownames_to_column("Barcode") 
```
Reduce to dichotomous variables (absence / presence)
```{r}
#Get df that has prevalence filtered pathway
#Dichotomise
poo_dicho <- hi %>% 
  mutate(across(where(is.numeric), ~ ifelse(. > 0, 1, 0)))
```
join functional data to metadata
```{r}
#Match samples with (filtered) tse
a <- as.data.frame(colData(tse))%>%
  tibble::rownames_to_column("Barcode")
rm(tse)

df <- dplyr::inner_join(poo_dicho, a, by = "Barcode")
```
Cox model for testing pathways associated with Incident CKD (outcome variable)
```{r}
##COX model
coxph_full_partial <- purrr::partial(survival::coxph,
                                formula = my_surv(HTN_AGEDIFF, INCIDENT_HTN) ~
                                  term + BL_AGE +
                                  MEN + BMI + PREVAL_DIAB +
                                  HFC + PREVAL_HFAIL_STRICT +
                                  CURR_SMOKE,
                                ties = "breslow") %>%
                      purrr::possibly()



#Take the column names from Pathway in shit dataframe as a list and mutate
try <- shit.10 %>% dplyr::select(Pathway) %>% 
  dplyr::mutate(results = purrr::map(Pathway, ~coxph_full_partial(data = df %>% dplyr::rename(term = .x)), .progress = TRUE))

try%>% 
  dplyr::filter(!is.na(results)) %>%
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, exponentiate = TRUE, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(term == "term") %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH")) %>%
  dplyr::arrange(p.value) %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 3)) %>%
  DT::datatable(caption = "Cox model for MetaCyc pathways")
```
```{r}
coxph_partial <- purrr::partial(survival::coxph,
                                formula = my_surv(HTN_AGEDIFF, INCIDENT_HTN) ~
                                  term + BL_AGE + MEN,
                                ties = "breslow") %>%
                       purrr::possibly()

try <- shit.10 %>% dplyr::select(Pathway) %>% 
  dplyr::mutate(results = purrr::map(Pathway, ~coxph_partial(data = df %>% dplyr::rename(term = .x)), .progress = TRUE))

try %>% 
  dplyr::filter(!is.na(results)) %>%
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, exponentiate = TRUE, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(term == "term") %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH")) %>%
  dplyr::arrange(p.value) %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 3)) %>%
  DT::datatable(caption = "Cox model for MetaCyc pathways")

```

Tranform to inverse_rank
```{r}
# Define the transformation function
transform_column <- function(x) {
  qnorm((rank(x, na.last = "keep") - 0.5) / sum(!is.na(x)))
}

# Apply the transformation to all columns
poo_rank <- hi
poo_rank[, -1] <- lapply(hi[, -1], transform_column)
```
Data wrangling to join functional data to metadata
```{r}
#Here you will have the correct number of samples
df <- dplyr::inner_join(poo_rank, a, by = "Barcode")
```
```{r}
##Cox for inverse rank
#Take the column names from Pathway in shit dataframe as a list and mutate
try <- shit.10 %>% dplyr::select(Pathway) %>% 
  dplyr::mutate(results = purrr::map(Pathway, ~coxph_full_partial(data = df %>% dplyr::rename(term = .x)), .progress = TRUE))

try%>% dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, exponentiate = TRUE, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(term == "term") %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH")) %>%
  dplyr::arrange(p.value) %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 3)) %>%
  DT::datatable(caption = "Cox model for MetaCyc pathways")
```
```{r}
try <- shit.10 %>% dplyr::select(Pathway) %>% 
  dplyr::mutate(results = purrr::map(Pathway, ~coxph_partial(data = df %>% dplyr::rename(term = .x)), .progress = TRUE))

try%>% dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, exponentiate = TRUE, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(term == "term") %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH")) %>%
  dplyr::arrange(p.value) %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 3)) %>%
  DT::datatable(caption = "Cox model for MetaCyc pathways")
```

#Remove any samples with zero
```{r}
#Take the column names from Pathway in shit dataframe as a list and mutate
#df <- shit.t %>% 
#  mutate_if(is.numeric, ~ifelse(abs(.) < 10^-10, NA, .)) %>%
 # dplyr::inner_join(a, by = "Barcode")

#try <- shit %>% 
#  dplyr::select(Pathway) %>% 
 # dplyr::mutate(results = purrr::map(Pathway, ~coxph_full_partial(data = df %>% dplyr::rename(term = .x)), .progress = TRUE, na.rm = TRUE))

#try%>% dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, exponentiate = TRUE, conf.int = TRUE)))  %>%
 # tidyr::unnest(results) %>%
 # dplyr::filter(term == "term") %>%
 # dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH")) %>%
 # dplyr::arrange(p.value) %>%
#  dplyr::mutate_if(is.numeric, ~round(.x, digits = 3)) %>%
 # DT::datatable(caption = "Cox model for MetaCyc pathways")
```


