---
title: "Incident Hypertension microbiome"
author: "Joonatan Palmu, Li-Fang Yeo"
date: "2025-06-30"
output: html_document
---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```
# Setup
```{r}
library(magrittr)
library(ANCOMBC)
library(mia)
library(scater)
```
```{r}
devtools::load_all()
```
# Command line arguments
```{r}
args <- list(
  optparse::make_option("--east", action="store_true", default=FALSE, help="Exclude Eastern Finland subpopulation [default \"%default\"]"),
  optparse::make_option("--west", action="store_true", default=FALSE, help="Exclude Eastern Finland subpopulation [default \"%default\"]"),
  optparse::make_option("--detection", type = "numeric", default = 0.1/100, help = "Detection limit [default %default]"),
optparse::make_option("--prevalence", type = "numeric", default = 5/100, help = "Prevalence limit [default %default]")) %>% 
  optparse::OptionParser(option_list = .) %>%
  optparse::parse_args()
```
```{r}
args %>% tibble::enframe(name = "Option", value = "Argument") %>% DT::datatable()
```
# Formatting options
```{r}
myggsave <- myggsavefactory()
```
```{r}
{ ggthemes::theme_tufte(base_family = "sans", base_size = 12) +
    ggplot2::theme(panel.border = ggplot2::element_rect(colour = "black", fill = NA),
                   axis.text = ggplot2::element_text(colour = "black", size = 10),
                   aspect.ratio = 1) } %>%
  ggplot2::theme_set()
```
# Data definitions
```{r}
#
vars <- list(BL_AGE = "Age",
             MEN = "Men",
             BMI = "BMI",
             PREVAL_DIAB = "Diabetes",
             SYSTM = "Systolic blood pressure",
             DIASM = "Diastolic blood pressure",
             HFC = "Healthy Food Choices",
             BP_TREAT = "Self-reported antihypertensive medication",
             PREVAL_HIBP = "Prevalent hypertension code",
             INCIDENT_HIBP = "Incident hypertension code",
             HIBP_AGEDIFF = "Hypertension code time",
             PREVAL_RX_RR_SPECIFIC = "Prevalent antihypertensive",
             INCIDENT_RX_RR_SPECIFIC = "Incident antihypertensive",
             RX_RR_SPECIFIC_AGEDIFF = "Antihypertensive agediff",
             PREVAL_HTN = "Prevalent hypertension",
             INCIDENT_HTN = "Incident hypertension",
             HTN_AGEDIFF = "Hypertension time (yrs)",
             CURR_SMOKE = "Smoking",
             EAST = "Eastern Finland",
             shannon = "Shannon diversity",
             observed = "Observed richness",
             total_reads = "Total reads", 
             PREVAL_CVD = "Prevalent cardiovascular disease") 
```
```{r}
tse <- readRDS("../../data/tse_mgs-20241118_104759.rds") %>%
  mia::transformAssay(assay.type = "counts", method = "relabundance") %>% 
  mia::addAlpha(assay.type = "counts", index = c("shannon", "observed"), name = c("shannon", "observed")) %>%
  tse_add_food_score(HFC) %>%
  tse_mutate(PREVAL_HTN = ifelse(PREVAL_HIBP + PREVAL_RX_RR_SPECIFIC > 0, 1, 0)) %>%
  tse_mutate(INCIDENT_HTN = ifelse(INCIDENT_HIBP + INCIDENT_RX_RR_SPECIFIC > 0, 1, 0)) %>%
  tse_mutate(HTN_AGEDIFF = pmin(HIBP_AGEDIFF, RX_RR_SPECIFIC_AGEDIFF)) %>% 
  tse_mutate(dplyr::across(c(MEN,
                             EAST,
                             BP_TREAT,
                             CURR_SMOKE,
                             dplyr::contains("INCIDENT"),
                             dplyr::contains("PREVAL")), as.factor)) %>%
  tse_filter(GRAVID %in% c(1, NA), BL_USE_RX_J01_1mo %in% c(0, NA)) %>% 
  tse_filter(SYSTM < 140, DIASM < 90, PREVAL_HTN == 0, BP_TREAT == 0) %>%
  tse_filter(dplyr::if_all(dplyr::one_of(names(vars)), not_na)) %>%
  tse_filter(total_reads > 50000) %>%
  tse_select(PREVAL_HTN, INCIDENT_HTN, HTN_AGEDIFF, names(vars))

```
# Characteristics
```{r}
tse %>%
  tse_meta(rownames = FALSE) %>%
  dplyr::mutate(INCIDENT_HTN = factor(ifelse(INCIDENT_HTN == 1, "Incident HTN" , "Without HTN"))) %>%
  mytableone(vars,fo =  ~ .| INCIDENT_HTN ) #%>%
  write.csv("tableone.csv")
```

# Alpha diversity
## Univariable adjusted for age and sex
```{r}
tse %>%
  tse_meta(rownames = FALSE) %>% 
  survival::coxph(my_surv(HTN_AGEDIFF, INCIDENT_HTN) ~ scale(shannon)+ scale(BL_AGE) + MEN,
                  data = .) %>% 
  broom::tidy(exponentiate = TRUE, conf.int = TRUE) %>%
  round_numeric_columns() %>% 
  DT::datatable(caption = "Alpha diversity")
```
```{r}
tse %>%
  tse_meta(rownames = FALSE) %>% 
  survival::coxph(my_surv(HTN_AGEDIFF, INCIDENT_HTN) ~ scale(observed)+ scale(BL_AGE) + MEN,
                  data = .) %>% 
  broom::tidy(exponentiate = TRUE, conf.int = TRUE) %>%
  round_numeric_columns() %>% 
  DT::datatable(caption = "Alpha diversity")
```
## Multivariable
```{r}
tse %>%
  tse_meta(rownames = FALSE) %>% 
  survival::coxph(my_surv(HTN_AGEDIFF, INCIDENT_HTN) ~
                    scale(shannon) + scale(BL_AGE) + MEN + scale(BMI) + PREVAL_DIAB + scale(HFC) +
                    PREVAL_CVD + CURR_SMOKE, data = .) %>% 
  broom::tidy(exponentiate = TRUE, conf.int = TRUE) %>%
  round_numeric_columns() %>% 
  #write.csv("cox-HTN.csv")
  DT::datatable(caption = "Alpha diversity")
```
```{r}
tse %>%
  tse_meta(rownames = FALSE) %>% 
  survival::coxph(my_surv(HTN_AGEDIFF, INCIDENT_HTN) ~
                    scale(observed) + scale(BL_AGE) + MEN + scale(BMI) + PREVAL_DIAB + scale(HFC) +
                    PREVAL_CVD + CURR_SMOKE, data = .) %>% 
  broom::tidy(exponentiate = TRUE, conf.int = TRUE) %>%
  round_numeric_columns() %>% 
  #write.csv("cox-HTN.csv")
  DT::datatable(caption = "Alpha diversity")

tse %>%
  tse_meta(rownames = FALSE) %>% 
  survival::coxph(my_surv(HTN_AGEDIFF, INCIDENT_HTN) ~
                    scale(observed) + scale(BL_AGE) + MEN + scale(BMI) + PREVAL_DIAB + scale(HFC) +
                    PREVAL_CVD + CURR_SMOKE, data = .) %>%
  survival::cox.zph() %>% survminer::ggcoxzph()
```
# Common taxa
## Agglomerate prevalent features
```{r}
subset_features_partial <-  purrr::partial(mia::agglomerateByPrevalence,
                                           x = tse,
                                           detection = args$detection,
                                           prevalence = args$prevalence,
                                           as_relative = TRUE) %>%
  purrr::possibly()
```
```{r}
taxa_subsets <- c("Species") %>%
  rlang::set_names() %>%
  purrr::map(subset_features_partial)
```
```{r}
melt_tse <- function(x, method = "clr") {
  mia::transformAssay(x, method = method, pseudocount = 1) %>% 
    mia::meltSE(add_row_data = TRUE, assay.type = method) %>%
    dplyr::mutate(FeatureID = glue::glue("GUT_{FeatureID}")) %>%
    dplyr::mutate(FeatureID = stringr::str_replace_all(FeatureID, c(" " = "_", "-" = "_"))) %>% 
    dplyr::select(SampleID, FeatureID, clr) %>%
    tidyr::spread(FeatureID, clr) %>% 
    dplyr::full_join(tse_meta(x), by = dplyr::join_by(SampleID == rownames))
}
```
```{r}
dfs <- taxa_subsets %>%
  purrr::map(melt_tse, .progress = TRUE)
```
# Cox's models
## Check available taxa
```{r}
#Generate a list of taxa, add the word GUT in front of the taxa
taxa_rank_list <- function(x, y) {
  list_names <- colnames(x) %>%
    stringr::str_subset("GUT_") %>%
    rlang::set_names()
}
```
```{r}
taxa_rank_list <- dfs %>%
  purrr::imap(taxa_rank_list, .progress = TRUE)
```
# Cox's model minimally adjusted
## age and sex
```{r}
coxph_partial <- purrr::partial(survival::coxph,
                                formula = my_surv(HTN_AGEDIFF, INCIDENT_HTN) ~
                                  taxon + BL_AGE + MEN,
                                ties = "breslow")
```
```{r}
cox_model_for_taxon <- function(rank, taxon) {
  dfs[[rank]] %>%
    dplyr::rename(taxon := {{taxon}}) %>% 
    coxph_partial(data = .)
}
```
```{r}
df_cox_results <- taxa_rank_list %>%
  purrr::map_df(~tibble::tibble(taxa = .x), .id = "rank") %>% 
  tidyr::gather(rank, taxa) %>%
  dplyr::mutate(results = purrr::map2(rank, taxa, cox_model_for_taxon, .progress = TRUE)) %>%
  dplyr::mutate(tidy = purrr::map(results, ~broom::tidy(.x, exponentiate = TRUE, conf.int = TRUE)))
```
```{r}
df_cox_results %>%
  dplyr::select(-results) %>% 
  tidyr::unnest(tidy) %>%
  dplyr::filter(stringr::str_detect(term, "taxon")) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH"))%>%
  dplyr::arrange(qval_fdr) %>%
  DT::datatable()
```
# Perform Cox's models
```{r}
coxph_partial <- purrr::partial(survival::coxph,
                                formula = my_surv(HTN_AGEDIFF, INCIDENT_HTN) ~
                                  taxon + BL_AGE +
                                  MEN + BMI + PREVAL_DIAB +
                                  HFC + PREVAL_CVD +
                                  CURR_SMOKE,
                                ties = "breslow")
```
```{r}
cox_model_for_taxon <- function(rank, taxon) {
  dfs[[rank]] %>%
    dplyr::rename(taxon := {{taxon}}) %>% 
    coxph_partial(data = .)
}
```
```{r}
df_cox_results <- taxa_rank_list %>%
  purrr::map_df(~tibble::tibble(taxa = .x), .id = "rank") %>% 
  tidyr::gather(rank, taxa) %>%
  dplyr::mutate(results = purrr::map2(rank, taxa, cox_model_for_taxon, .progress = TRUE)) %>%
  dplyr::mutate(tidy = purrr::map(results, ~broom::tidy(.x, exponentiate = TRUE, conf.int = TRUE))) 
```
```{r}
df_cox_results %>%
  dplyr::select(-results) %>% 
  tidyr::unnest(tidy) %>%
  dplyr::filter(stringr::str_detect(term, "taxon")) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH"))%>%
  dplyr::arrange(qval_fdr) %>%
  DT::datatable()
```
# Check PH assumptions
```{r}
zph_validation <- function(x){
  survival::cox.zph(x)$table %>% tibble::as_tibble(rownames = "term")
}
```
```{r}
df_cox_results %>%
  dplyr::mutate(results = purrr::map(results, zph_validation, .progress = TRUE)) %>% 
  tidyr::unnest(results) %>%
  dplyr::arrange(p) %>%
  DT::datatable()
```
# Kaplan Meier
```{r}
kaplan_meier_plot <- function(rank, taxon, n = 4) {
  dfs[[rank]] %>%
    dplyr::rename(taxon := {{taxon}}) %>%
    dplyr::mutate(taxon_tile = dplyr::ntile(taxon, n)) %>% 
    ggsurvfit::survfit2(my_surv(HTN_AGEDIFF, INCIDENT_HTN) ~ taxon_tile,
                        data = .) %>%
    ggsurvfit::ggsurvfit(linewidth = 1) +
    #ggsurvfit::add_confidence_interval() +
    ggsurvfit::add_risktable(risktable_stats = "{cum.event}") +
    ggsurvfit::add_risktable_strata_symbol(symbol = "\U25CF", size = 10) +
    ggplot2::labs(title = taxon) +
    ggplot2::scale_color_grey(start = 0.9, end = 0)
}

```
```{r}
df_cox_results %>%
  tidyr::unnest(tidy) %>%
  dplyr::filter(stringr::str_detect(term, "taxon")) %>%
  dplyr::filter(p.value < 0.05) %>%
  dplyr::arrange(p.value) %>% 
  dplyr::mutate(plot = purrr::map2(rank, taxa, kaplan_meier_plot)) %>%
  dplyr::pull(plot) %>% 
  gridExtra::arrangeGrob(grobs = ., ncol = 3) %>%
  myggsave("km_hypertension", width = 15, height = 30)

```
# ANCOM-BC2
```{r}
tse_species_BC <- mia::agglomerateByPrevalence(tse, rank = "Species",
                                              detection = 0.1/100,
                                              prevalence = 5/100,
                                              as_relative = TRUE)
```
```{r}
pcol <- glue::glue("q_INCIDENT_HTN1")
```

```{r}
ancombc2_species <- ANCOMBC::ancombc2(data = tse_species_BC,
            fix_formula = "INCIDENT_HTN + BL_AGE + MEN + BMI + PREVAL_DIAB + HFC + PREVAL_CVD + CURR_SMOKE",
                         assay.type = "counts",
                         p_adj_method = "fdr",
                         verbose = TRUE,
                         neg_lb = TRUE)
```
```{r}
df <- ancombc2_species$res %>%
  dplyr::select(taxon, contains("INCIDENT_HTN")) %>%
  dplyr::arrange(!!rlang::sym(pcol))
```
```{r}
df %>%
  round_numeric_columns() %>% 
  DT::datatable(caption = "INCIDENT")

df %>%
  round_numeric_columns() %>% 
  write.csv("ancombc-full.csv")
```
```{r}
ancombc2_species <- ANCOMBC::ancombc2(data = tse_species_BC,
            fix_formula = "INCIDENT_HTN + BL_AGE + MEN",
                         assay.type = "counts",
                         p_adj_method = "fdr",
                         verbose = TRUE,
                         neg_lb = TRUE)
```
```{r}
df <- ancombc2_species$res %>%
  dplyr::select(taxon, contains("INCIDENT_HTN")) %>%
  dplyr::arrange(!!rlang::sym(pcol))
```
```{r}
df %>%
  round_numeric_columns() %>% 
  DT::datatable(caption = "INCIDENT")

df %>%
  round_numeric_columns() %>% 
  write.csv("Ancombc-mini.csv")
```
PERMANOVA
```{r}
tse<- tse %>%  mia::mergeFeaturesByRank(rank = "Species") %>% 
    mia::transformAssay(method = "relabundance") %>%
    scater::runMDS(FUN = getDissimilarity,
                assay.type = "relabundance",
                method = "bray",
                na.action = na.exclude,
                name = "MDS_bray")

# Create ggplot object
p <- plotReducedDim(tse, "MDS_bray", colour_by = "INCIDENT_HTN")

# Calculate explained variance
e <- attr(reducedDim(tse, "MDS_bray"), "eig")
rel_eig <- e / sum(e[e > 0])

# Add explained variance for each axis
p <- p + labs(
    x = paste("PCoA 1 (", round(100 * rel_eig[[1]], 1), "%", ")", sep = ""),
    y = paste("PCoA 2 (", round(100 * rel_eig[[2]], 1), "%", ")", sep = "")
    )

p +
  ggplot2::scale_colour_manual(name = "Incident hypertension",
                               labels = c("0" = "Normotensive",
                                          "1" = "Hypertensive"),
                               values = c("0" = "darkslategrey",
                                          "1" = "orange")) +
  ggplot2::theme_classic() +
  ggplot2::theme(
  legend.title = element_text(size = 16),
  legend.text = element_text(size = 14),
  axis.text = element_text(size = 14),
  axis.title = element_text(size = 16))
```
```{r}
#Permanova
beta <- tse %>%  mia::mergeFeaturesByRank(rank = "Species") %>% 
    mia::transformAssay(method = "relabundance") %>%
    mia::runRDA(assay.type = "relabundance",
                formula = assay ~ BL_AGE + MEN + INCIDENT_HTN,
                distance = "bray",
                na.action = na.exclude)
# Store results of PERMANOVA test
rda_info <- attr(SingleCellExperiment::reducedDim(beta, "RDA"), "significance")

rda_info$permanova |>
  knitr::kable()
```
```{r}
rda_info$homogeneity |>
  knitr::kable()
```
```{r}
{ miaViz::plotRDA(beta,
                "RDA",
                add.vectors = FALSE,
                add.ellipse = "colour",
                colour_by = "INCIDENT_HTN") +
    ggplot2::scale_colour_manual(name = "Incident hypertension",
                                 labels = c("0" = "Normotensive",
                                            "1" = "Hypertensive"),
                                 values = c("0" = "gray",
                                            "1" = "black")) } %>% 
  myggsave("hypertension_rda", width = 10, height = 10)
```
Full model
```{r}
beta2 <- tse %>%  mia::mergeFeaturesByRank(rank = "Species") %>% 
    mia::transformAssay(method = "relabundance") %>%
    mia::runRDA(assay.type = "relabundance",
                formula = assay ~ BL_AGE + MEN + BMI + PREVAL_DIAB + HFC +
                  PREVAL_CVD + CURR_SMOKE + INCIDENT_HTN,
                distance = "bray",
                na.action = na.exclude)
```
```{r}
# Store results of PERMANOVA test
rda_info <- attr(SingleCellExperiment::reducedDim(beta2, "RDA"), "significance")
```
```{r}
rda_info$permanova |>
  knitr::kable() %>%
  write.csv("./permanova-HTN-full.csv")
```

```{r}
rda_info$homogeneity |>
  knitr::kable() %>%
  write.csv("homogeneity-HTN-full.csv")

saveRDS(beta2, "beta-HTN-full.rds")
```

```{r}
{ miaViz::plotRDA(beta2,
                "RDA",
                add.vectors = FALSE,
                add.ellipse = "colour",
                colour_by = "INCIDENT_HTN") +
    ggplot2::scale_colour_manual(name = "Incident hypertension",
                                 labels = c("0" = "Normotensive",
                                            "1" = "Hypertensive"),
                                 values = c("0" = "gray",
                                            "1" = "black")) } %>% 
  myggsave("hypertension_rda", width = 10, height = 10)
```
